---
title: "class05: data viz with ggplot"
author: "Seona Patel (PID: A69035519)"
format: pdf
---

Today we are playing with plotting and graphics in R.

There are lots of ways to make cool figures in R.
There is "base" R graphics (`plot()`, `hist()`, `boxplot()`, etc.)

There is also add-on packages, like **ggplot**

```{r}
head(cars, 3)
```

Let's plot this with "base" R: 

```{r}
plot(cars)
```
```{r}
mtcars
```

```{r}
head(mtcars)

```
Let's plot `mpg` vs `disp`

```{r}
plot(mtcars$mpg, mtcars$disp, pch=16)
```

```{r}
hist(mtcars$mpg)
```

## GGPLOT
The main function in the ggplot2 package is **ggplot()** 
first I need to install the **ggplot2** package. I can install any package with the function`install.packages()`
> **N.B.** Never want to run `install.packages()` in my quarto document because it will install it every time you render the page. Instead, install it in the console so you only have to do it once. 

```{r}
library(ggplot2)
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point()

```

```{r}
ggplot(cars) +
  aes(speed) +
  geom_histogram()
```

Every ggplot needs at least 3 things:

- The **data** (given with `ggplot(cars)`)
- The **aesthetic** mapping (given with `aes()`)
- The **geom** (given by `geom_point()`)

> For simple canned graphs "base" R is nearly always faster (in writing code and running)

### Adding more layers

Let's add a line and a title, subtitle, and caption as well as custom axis labels

```{r}
library(ggplot2)
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_line()

```

```{r}
library(ggplot2)
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth(method = 'lm', se=FALSE)

```

```{r}
library(ggplot2)
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth(method = 'lm', se=FALSE) + 
  labs(title = "Silly Plot", x= "Speed (MPH)", y="Distance (ft)", caption = "when is tea time anyway??") +
  theme_bw()
```

##Plot some expression data

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```
> Q1. How many genes are in this wee dataset?

There are `r nrow(genes)` rows in this dataset

> Q2. How many columns are there are what are their names?

There are `r ncol(genes)` columns in this dataset
```{r}
colnames(genes)
```
The column names are `r colnames(genes)` 

> Q3. How many "up" regulated genes are there?

There are `r sum(genes$State == 'up')` UP genes

```{r}
table(genes$State)
```
> Q4 . What fraction of total genes is up-regulated in this dataset?

```{r}
round( table(genes$State)/nrow(genes) * 100, 2 )
```


```{r}
p <- ggplot(genes) + 
    aes(x=Condition1, y=Condition2, col=State) +
    geom_point() +
  scale_color_manual(values=c("blue", "gray", "red"))
```
```{r}
p + labs(title='look at me')
```

Silly exmaple of adding labels
```{r}
library('ggrepel')
```

```{r}
ggplot(genes) +  
  aes(x=Condition1, y=Condition2, col=State, label=Gene) +
  geom_text_repel(max.overlaps = 60)+
  geom_point() +
  scale_color_manual(values=c("blue", "gray", "red")) +
  theme_bw()
```
## Going Further

Playing with some different layers and the gapminder dataset

```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
head(gapminder)
```
```{r}
tail(gapminder)
```
A first plot 
alpha in geom_point makes the points more transparent (useful when you have a lot of overlapping datapoints)
```{r}
ggplot(gapminder)+
  aes(gdpPercap, lifeExp, col=continent, size=pop) +
  geom_point(alpha =0.2)+
  facet_wrap(~continent) +
  theme_bw()
```
```{r}
library(gapminder)
library(gganimate)

# Setup nice regular ggplot of the gapminder data
ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, colour = country)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  # Facet by continent
  facet_wrap(~continent) +
  # Here comes the gganimate specific bits
  labs(title = 'Year: {frame_time}', x = 'GDP per capita', y = 'life expectancy') +
  transition_time(year) +
  shadow_wake(wake_length = 0.1, alpha = FALSE)
```

```{r}
library(patchwork)

# Setup some example plots 
p1 <- ggplot(mtcars) + geom_point(aes(mpg, disp))
p2 <- ggplot(mtcars) + geom_boxplot(aes(gear, disp, group = gear))
p3 <- ggplot(mtcars) + geom_smooth(aes(disp, qsec))
p4 <- ggplot(mtcars) + geom_bar(aes(carb))

# Use patchwork to combine them here:
(p1 | p2 | p3) /
      p4
```

