---
title: "Class 10: Halloween Mini-Project"
author: "Seona Patel (A69035519)"
format: pdf
---
# Background

Is it Halloween today and to help get in the mood for tonight we will explore some candy preference data from 538.

## 1. Importing the candy file

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```
```{r}
flextable::flextable(head(candy))
```

We will take a whirlwind tour of this dataset and in the process answer the questions highlighted in red throught this page that aim to guide your exploration process. We will then wrap up by trying Principal Component Analysis (PCA) on this dataset to get yet more experience with this important multivariate method. It will yield a kind of “Map of Hallowen Candy Space”. How cool is that! Let’s explore…

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

```{r, message=FALSE}
library(dplyr)
candy |> nrow()
```

There are 85 different candy types in this dataset.

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity ==1)

```

There are 38 fruity candy types in the dataset.

 ## 2. What is your favorite candy?

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

My favorite candy is M&M's and it's winpercent is 66.57458%

```{r}
candy["M&M's", ]$winpercent

```


> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```
The winpercent value for "Kit kat" is 76.7686%.

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

The winpercent value for Tootsie Roll Snack Bars is 49.6535%


```{r}
library("skimr")
skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

winpercent is on a 0-100 scale rather than 0-1 which is what the rest of the variables are on. 

>Q7. What do you think a zero and one represent for the candy$chocolate column?
Hint: look at the “Variable type” print out from the skim() function. Most varables (i.e. columns) are on the zero to one scale but not all. Some columns such as chocolate are exclusively either zero or one values.

0 and 1 are logicals, so 0 means it is not a chocolate type and 1 means it is a chocolate type. 


> Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)
ggplot(candy,)+
  aes(winpercent) +
  geom_histogram(bins=20)
```

> Q9. Is the distribution of winpercent values symmetrical?

No, the distribution looks right skewed rather than symmetrical.

```{r}
ggplot(candy,)+
  aes(winpercent) +
  geom_density()
```


> Q10. Is the center of the distribution above or below 50%?

It depends on how you measure centrality, if you use the mean it is above 50% but if you use median it is below 50%

```{r}
mean(candy$winpercent)
```

```{r}
summary(candy$winpercent)

```

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}

# 1. Find all chocolate candy in the dataset
choc.inds <- as.logical(candy$chocolate)
choc.candy <- candy[choc.inds,]
# 2. Extract their `winpercent` values
choc.win <- choc.candy$winpercent
# 3. Find the mean of these values 
choc.mean <- mean(choc.win)
# 4-6. DO the same for fruity candy

fruity.inds <- as.logical(candy$fruity)
fruity.candy <- candy[fruity.inds,]
fruity.wins <- fruity.candy$winpercent
fruity.mean <- mean(fruity.wins)
# Which mean value is higher 

choc.mean
fruity.mean


```
Chocolate candy is ranked higher than fruity candy on average.

> Q12. Is this difference statistically significant?

```{r}
t.test(choc.win, fruity.wins)

```
Yes this is statistically significant.

## 3. Overall Candy Rankings

Let’s use the base R order() function together with head() to sort the whole dataset by winpercent. Or if you have been getting into the tidyverse and the dplyr package you can use the arrange() function together with head() to do the same thing and answer the following questions:

> Q13. What are the five least liked candy types in this set?

```{r}
candy |> arrange(winpercent) |> head(5)
```
The least liked candy types in this set are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters

> Q14. What are the top 5 all time favorite candy types out of this set?
Hint: Using base R we could use head(candy[order(candy$winpercent),], n=5), whilst using dplyr we have: candy %>% arrange(winpercent) %>% head(5). Which apprach do you prefer and why?

```{r}
candy |> arrange(desc(winpercent)) |> head(5)
```
The top 5 favorite candy types are Reese's Peanut Butter cup, Reese's Miniatures, Twix, Kit Kat, And Snickers


To examine more of the dataset in this vain we can make a barplot to visualize the overall rankings. We will use an iterative approach to building a useful visulization by getting a rough starting plot and then refining and adding useful details in a stepwise process.

> Q15. Make a first barplot of candy ranking based on winpercent values.
HINT: Use the aes(winpercent, rownames(candy)) for your first ggplot like so

```{r}
ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?
HINT: You can use aes(winpercent, reorder(rownames(candy),winpercent)) to improve your plot.

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()
```

Time to add some useful color
Let’s setup a color vector (that signifies candy type) that we can then use for some future plots. We start by making a vector of all black values (one for each candy). Then we overwrite chocolate (for chocolate candy), brown (for candy bars) and red (for fruity candy) values.


```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

Now let’s try our barplot with these colors. Note that we use fill=my_cols for geom_col(). Experement to see what happens if you use col=mycols.

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

Now, for the first time, using this plot we can answer questions like:
> Q17. What is the worst ranked chocolate candy?

The worst ranked chocolate is Sixlets.

> Q18. What is the best ranked fruity candy?

The best ranked fruity candy is Starburst.



## 4. Taking a look at pricepercent

What about value for money? What is the the best candy for the least money? One way to get at this would be to make a plot of winpercent vs the pricepercent variable. The pricepercent variable records the percentile rank of the candy’s price against all the other candies in the dataset. Lower vales are less expensive and high values more expensive.

To this plot we will add text labels so we can more easily identify a given candy. There is a regular geom_label() that comes with ggplot2. However, as there are quite a few candys in our dataset lots of these labels will be overlapping and hard to read. To help with this we can use the geom_text_repel() function from the ggrepel package.

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
ord <- order(candy$pricepercent, decreasing = FALSE)
head( candy[ord,c(11,12)], n=5 )

```
Of the least expensive candies, Tootsie Roll Midgies has the highest winpercent, making it the greatest bang for your buck. 

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?
Hint: To see which candy is the most expensive (and which is the least expensive) we can order() the dataset by pricepercent.

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```
The top 5 most expensive candies are Nik L Nip, Nestle Smarties, Ring pop, Hershey's Krackel, and Hershey's Milk Chocolate. Of these, the least popular is Nik L Nip. 

> Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_col()
```


```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```

# Correlation

```{r}
cor(candy)
```
```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)

```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values

Fruity and chocolate are anti-correlated.

> Q23. Similarly, what two variables are most positively correlated?

winpercent and chocolate are most strongly positively correlated.


## 6. Principal Component Analysis

Let’s apply PCA using the prcomp() function to our candy dataset remembering to set the scale=TRUE argument.

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

Now we can plot our main PCA score plot of PC1 vs PC2.


```{r}
plot(pca$x[,1:2])
```

```{r}
ggplot(pca$x) + 
  aes(PC1, PC2, label=rownames(pca$x)) +
  geom_point(col=my_cols) +
  geom_text_repel(col=my_cols)+
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

Dont forget about your variable "loadings" - how the original variables contribute to your new PC's...

```{r}
ggplot(pca$rotation) +
  aes(PC1, rownames(pca$rotation)) +
  geom_col()
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?
HINT. pluribus means the candy comes in a bag or box of multiple candies.

The variables fruity, hard, and pluribus are picked up strongly by PC1 in the positive direction. his makes sense as the fruity candy tends to be hard and come in a  bag or box of multiple candies.
